Distribution-oblivious Programming with Dish

### 2019-11-19 pldi/draft.tex
Draft content for intermediate sections, to estimate budget and feasibility

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}

@@

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
\newcommand{\review}[1]{{\color{red}#1}}

@@

Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or can process data that would not fit into any single machine.
% offer notable benefits over their centralized counterparts:
%   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
% Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
% The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
  % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.

To understand this sharp contrast, consider a Bash pipeline calculating term frequencies over a set of inputs:

@@

\noindent
Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, reduces duplicates to a counter, sorts counters in descending order, picks the top five results, and writes them to a file \ttt{out}.
Combining several features, \unix makes small tasks easy to express;
  for a user with \emph{one} computer and small input size, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
% composition allows the composition of general primitives

@@

Unfortunately, developing $p_1$'s distributed equivalent requires a significant effort.
% For a user with many computers and larger inputs, 
% The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
For simple pipelines, fitting into restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
More complex pipelines, such as the ones presented in later sections, would involve a full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}. %  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
In both cases, manual rewriting is expensive and can introduce new bugs, cascading changes, and divergence from legacy functionality.
Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?

@@

To answer this question, this paper presents \sys, a shell \todo{preprocessor} that transforms shell pipelines into their distributed equivalents enabling \emph{distribution-oblivious programming}.
The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required to distribute a computation.
\sys builds on this insight with a careful study of the distributability properties of shell primitives and commands, a pipeline rewriting pass that identifies maximal sub-expressions as candidates for scale-out, and a late-bound command-prefixing scheme for deferring scheduling decisions to the distributed planner (\ie when runtime information is available).
A \sys-enabled $p_1$ will run \ttt{tr} and \ttt{uniq} in completely parallel streams, use a mostly-parallel tree of \ttt{sort}s, run \ttt{head} in one of the streams, and (trivially) merge streams in \ttt{out}.

The combination of automated transformations for program distribution with the ability to maintain correct non-distributed semantics results in several benefits.
% \sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
First, shell users composing pipelines (or simply running legacy pipelines on massive datasets) can see scalability benefits without any manual effort---no need for \ttt{qsub}, SLURM, calls to \ttt{parallel}, or any manual rewriting.
Second, developers of new shell commands can use a concise a domain-specific language to express their distributability properties rather than using ad-hoc, command-specific flags such as {\tt t}  \#threads, \ttt{-t}, \ttt{-p} \etc
Most importantly, \sys provides an architectural lesson for system designers---namely that large-scale research efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.

@@

@@

>>>>

\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{amssymb}
\usepackage{soul}
\usepackage{xspace}
\usepackage{color}
\usepackage{xcolor}
\usepackage{upquote}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{cleveref}

@@

% For comments
\newcommand{\eat}[1]{}
\newcommand{\TODO}[1]{\hl{\textbf{TODO:} #1}\xspace}
\newcommand{\todo}[1]{\hl{#1}\xspace}
\newcommand{\nv}[1]{[{\color{cyan}#1 --- Nikos}]}
\newcommand{\kk}[1]{[{\color{magenta}#1 --- KK}]}
\newcommand{\review}[1]{{\color{red}#1}}

@@

Distributed systems offer significant benefits over their centralized counterparts---for example, they can speed up expensive computations or process data that would not fit into any single machine.
% offer notable benefits over their centralized counterparts:
%   using partitioning and replication, they can process and store data with increased throughput and fault-tolerance.
% Yet, only a minority of developers, employed by the select few companies that deal with massive datasets, have the luxury of engineering software systems with distribution baked in from the start.
% The remaining majority starts by developing and deploying software in a centralized manner---that is, \emph{until} there is a significant change of requirements, such as a load increase.
Despite these benefits, their development remains different from and significantly more difficult than their centralized counterparts.
Whereas anyone can quickly stitch together a Bash script to compute on a single computer, 
  % domain-experts routinely glue scripts together to process and share data. % without the help of a computing expert.
   scaling out to multiple ones requires expert labor around ``point'' solutions with expensive setups, restricted programming interfaces, and exorbitant composition costs~\cite{taurus:14, dios:13, andromeda:15, pywren:17, futuredata:18, nefele:18}.

To understand this sharp contrast, consider calculating term frequencies over a set of input files:

@@

\noindent
Program $p_1$ creates a character stream, breaks it into words, transliterates to lower-case, sorts to group duplicates, summarizes them with a counter, sorts them in descending order, picks the top five results and writes them to a file \ttt{out}.
Combining several features, \unix makes small tasks easy to express;
  for a user with \emph{one} computer and small input, this pipeline takes a few seconds to compose and execute~\cite{bentley1986literate}.
% composition allows the composition of general primitives

@@

For a user with many computers and larger inputs, however, developing $p_1$'s distributed equivalent requires a significant effort.
% The scope of such rewrites, and therefore the cost of manual effort, can vary considerably.
For simple pipelines that capture in restricted models of computation, this effort amounts to expressing the computation using the primitives provided by a big-data framework~\cite{mapreduce:08, ciel:11, spark:12, naiad:13} or domain-specific language~\cite{alvaro2011consistency, distal:13, meiklejohn2015lasp}---an unjustifiable cost for one-off pipelines that take a few minutes to compose (but are applied to large datasets).
More complex pipelines, such as the ones presented in later sections, require full-fledged distributed programming language~\cite{erlang:96, lopes1997d, acute:05, mace:07, cloudhaskell:11, ScalaLoci:18}. %  or a distributed operating system---\eg Plan9's \ttt{rc} shell.
In these cases, manual rewriting is expensive and can introduce new bugs, cascading changes, or divergence from legacy functionality.
Could the generation and execution of $p_1$'s distributed version be fully and correctly automated?

@@

The key insight behind \sys is that the language of the Unix shell already encodes stream processing, providing most of the information required for distributing a computation.
\sys builds on this insight with a careful study of the scalability properties of shell primitives and its commands, paired with a novel rewriting scheme that identifies maximal pipeline sub-expressions that are candidates for scale-out.
We term the result \emph{distribution-oblivious programming}---a series of techniques that fully automates program distribution while provably maintaining correct (non-distributed) semantics.

Key benefits include the ability to scale legacy computations onto
\sys converts legacy shell pipelines into their distributed equivalents fully automatically, offering $100\times$ improvements in performance without a single line of additional code or annotation.
Most importantly, \sys provides an architectural lesson:
  the current design shows that large-scale efforts in the distributed- and operating-system literature to provide a \unix-like distributed equivalent could have been simplified by a thin (but sophisticated) rewriting shim like the one \sys provides.

@@

\kk{We have to mention character vs line based parallelism somewhere,
  and that we choose to see lines as the data quantum}

@@

\section{Extensibility}

ULTRA DRAFTY:

\kk{Text point: A problem that is specific to the shell is
  extensibility.}

The major characteristic that makes the shell popular and very useful
is its extensibility, as also mentioned in \Cref{bg}. Commands change,
get extended, and each user continuously installs more commands to use
in their system.

\kk{Text point: This restricts the space of solutions for distribution
  of shell scripts}

This means that any one-time solution is not satisfactory, since it
would very easily become obsolete, or it would only handle a limited
amount of shell scripts.

\kk{Text point: The characteristics of a good solution}

A solution that applies to shell programms should thus be extensible,
in the sense that adding or changing a command, doesn't invalidate the
assumptions that the tool makes. In our case, the information about
the command categories. A developer of a new command, should be able
to categorize it accordingly, and succinctly describe how the system
can distribute a command.

\kk{Solution: A lightweight language that can be used by a command
  developer to categorize it for distribution.}

We address this issue by designing a lightweight language whose
purpose is to categorize a command, describes how it reads its input
(meaning in what order), and how to distribute it (in case it is
pure). \kk{I am not sure if the order in which a command reads its
  input will be part of the language since many commands can be made
  to read from one place (like stdin or their first argument). Even in
  that case though, where the command only reads from one argument, we
  have to make sure that there is a way for the user to pinpoint which
  argument does a command read from.}

\subsection{Categorization Language}

\kk{Text point: Introduce problem: Most shell commands, belong to
  different classes depending on their flags, arguments}

Consider the command \texttt{cat} \kk{Maybe that is not he best
  example. TODO: Find a good example of a simple command that is in
  different classes depending on the arguments.}. When used without
any flag it is stateless, etc...

\kk{Text point: Our language must be able to categorize commands to
  different classes according to their arguments.}

\kk{We should have a crisp point about why we have this categorization
  language, and why don't we just allow someone to write a function in
  python for each command, that given the command and its arguments,
  returns whether the command category. Possible arguments include,
  the fact that this language is simpler to use, especially by non
  experts that just run some script but have installed some commands
  that are not ``supported''. Another possible argument is that we
  could be able to reason about the constructs of the categorization
  lagnauge, and that they will be easier to read and
  understand. Another argument is that since these categorizations
  should be shareable among users, it would be bad to execute
  arbitrary python code, so using this language, expressivity is
  limited. All of these arguments are a little bit weak though. Niko,
  what do you think?}

Sketch of the language:

Some kind of lightweight annotation configuration language (like
yaml), that contains a record for each command.

A record contains a sequence of predicates => categories. Each
predicate is a combination of not, or, and and the atomic predicates
are the presence of a flag, or an equality that checks that the value
of a flag is equal to something (because flags can also have
values). The final predicate can be left empty, and it is trivially true.

Interpreter:

Given a command and the set of its arguments, the intepreter returns
the command category. In does so by returning the category of the first
predicate that matches.

\kk{Would it be meaningful to give the syntaxx of the language
  formally?}

\kk{If we talk about the language it would be good to give some
  statistics about how many commands (out of the ones that we have
  classified) can be represented with one, two clauses etc. If we have
  a different solution (like python function that given a command and
  its arguments returns the category) then we could talk about how
  many lines of code it took to categorize all the commands that we
  did.}

As described in distributability\ref{}, the abstract notion of a
program that we have is programs with one input and one output
stream. The command developer must be able to express the input
argument of a commmand (or if it is the stdin).

\kk{Question: Do we need to handle all commands (even the ones that
  could have their input split in several different arguments? or can
  we somehow easily make these ones just take their input from one
  place (e.g. by adding a cat before them).}

\kk{Possible solutions for this (if we have already turned commands
  with more that one input to commands with one input): Have a
  positional reference to the argument that is input (or ``stdin'' if
  it is the stdin). Example: the last argument [-1] of grep is its
  input.}

\kk{It seems that even the arguments might depend on flags (e.g. grep
  checks the current directory if asked recursively, or the stdin
  otherwise), so maybe we should again have a sequence of predicates}

\kk{Note: Since a categorization language will probably not be
  complete (especially the one showing the input arguments) there
  should be a backup mechanism, when the language is not expressive
  enough, like a python function that categorizes the command, and
  returns its input argument if it is stateless.}

\subsection{Pure Commands}

Even though stateless commands are a big percentage of all of them,
there are a lot of useful and popular pure commands.

As mentioned in distributability, pure commands can be broken down
into different categories. We might be able to get ideas about this
from this paper:
\url{http://www.cs.toronto.edu/~azadeh/papers/pldi17-ex.pdf} and its
continuation in PLDI 2019.

Can we find a solution for the commands in coreutils?

\section{Front End}

We parse a shell script using a posix compliant parser that returns
the ast.

\kk{Text point: The AST Is not a very good data structure to
  manipulate for distribution, so we translate it to a kind of a
  dataflow graph.}

\subsection{Distributable Subtrees}

\kk{Text point: Problem: Not all parts of a shell script are
  distributable, so we cannot arbitrarily distribute}

Even though commands are split in categories, and several pipelines
are distributable, not all parts of a shell script are stateless and
several of them have effects that could affect the behaviour of a
program after distribution.

\kk{Solution: We have found that distributing subtrees of the shell
  AST is safe. Can we have a semi-formal argument about that?}

Mention all the different constructs that our front end handles, and
why it is safe to add them to the distribution pipeline.

\subsection{Intermediate Representation/Dataflow graph}

Mention its constructs, and its semantics, and why it is a bit
different that standard dataflow graphs (where usually incoming edges
are either consumed with some arbitrary interleaving, or pair by
pair).

\kk{Text point: The main shell abstraction is files, which are batches
  or streams of characters (or lines) where the order matters. So we
  want a representation that captures order of inputs and outputs (and
  ours does because its node in the graph has a sequence of inputs.}

Explain that the intermediate representation represents workers that
communicate through files, and that it captures parallelism both in
the operator level, as well as data parallelism.

\section{Distribution Planner}

\kk{Maybe this should be merged with the one above, especially if it
  talks about how distribution happens on the intermediate
  representation.}

\subsection{Scaling up the distributed computation/Vertically expanding the graph}

The planner acts on the dataflow graph, by first expanding it ``as
much as possible'' to expose parallelization opportunities. It uses
knowledge about the inputs of each stateless and pure command.

\kk{Explain the algorithm and how it pushes parallelization through
  the graph, while preserving correctness.}

\kk{Give an example}

\subsection{Mapping operators to nodes}

Explain the simple algorithm that we used to minimize intra node
communication, that tries to map contiguous parts of the graph to the
same node. We could do this by having an algorithm that minimizes the
number of cuts or something.

We don't try to make a proper planner, as there is a lot of work on
operator placement etc. that we can borrow from.

\subsection{Just in time planning}

Important point: Shell is extremely dynamic. Because of this, a
distributor cannot decide how to distribute a subexpression
statically, by just reading the script, (as in many other systems like
MapReduce, Spark, etc), since most of the information is not there
before the script starts executing. Environment variables,
unexpanded(unevaluated) strings \kk{Make sure that terminology is
  consistent with Greenberg}, could all contain information that is
valuable to make the distribution plan \kk{give an example}.

Future work on this: calling the shell shtepper to only partially
evaluate strings, just expand arguments, and not do any significant
computation.

\section{Limitations}

\kk{This is not really a section on its own, but rather part of something else.}

\begin{itemize}
\item Fault Tolerance
\item Distributed Planning: The planner at the moment is naive, and
  there is a huge literature on operator placement of distributed
  dataflow graphs that can be used to improve it.
\item Support of shell constructs: At the moment we have been
  conservative in the shell subsets that we handle, so that we don't
  introduce any unsafe parallelism. The point of this work was not to
  be able to handle a complete distributable subset of the shell, but
  rather a significantly big part of it, to demonstrate that there can
  be performance benefit from it.
\item Support distributing different classes: At the moment we are
  only dealing with stateless and pure commands, that is the ones that
  are most popular but also can offer the biggest benefits from
  parallelizing them. However, it would be interesting to explore the
  distributability of the file system commands, by using a distributed
  file system. THis would also require considerations about
  replication, consistency, and failures.
\end{itemize}
